Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-07-13 extended_catalog/paper/extended_catalog.tex
more revisions

\pointlike also bins $\gamma$-rays depending upon whether they
convert in the front or back parts of the detector. Since the PSF
is smaller for front entering photons, binning each conversion type
independently improves the tool's sensitivity.
Furthermore, \pointlike uses a sparse matrix representation of the spatial
bins where only the bins with counts in them are used. \pointlike
then independently integrates the overall model predicted counts in the region.

The intelligent binning and sparse matrix representation of the data
effectively interpolates between a binned analysis at low energy (where
the resolution is poor and there are many counts) and an unbinned
analysis at high energy (where the resolution is good but there are
few counts). This approach provides a significant savings in time while
introducing small numerical error.

Finally, \pointlike is written mostly in the \python programming
language and uses {\em numpy} and {\em scipy} to vectorize most of the
calculations. This provides a reasonable trade off between computational
efficiency and ease of development. By leaving the interface as a series
of \python modules, advanced functions including extension fitting
(described in section~\ref{extesion_fitting}) and a dual localization
(described in section~\ref{dual_localization_method}) could be quickly
added on top of the core functionality.

@@

We modified \pointlike to allow modeling the spatial extension of
of a source. It is assumed that the spatial and spectral information
completely decouple and that the spatial shape can be represented by a
simple geometric function.  The code consistently convolves the extended
source shape with the PSF (as a function of energy) when computing the
model predicted counts in a bin.

\pointlike uses the \minuit fitting library to simultaneously maximize
the likelihood by fitting the position and extension of an extended
source (\cite{minuit_documentation}).  As will be described in
section~\ref{monte_carlo_validation}, the statistical significance
of the detection of an extended source can only be correctly calculated
when all of the parameters of a model are simultaneously fit

For each position and extension,
the spectral parameters of the sky model are refit. To avoid projection
effects, what is directly fit is not the longitude and latitude of
the source but instead the displacement in a rotated reference frame.
The significance of the extension of a source can be calculated from
\tsext, which is defined as twice the increase in log likelihood between
the extended source model at its best fit values and the
point source model at its best fit values:
\begin{equation}
  \tsext=2\log(LL_\text{ext}/LL_\text{ps}).
\end{equation}
The interpretation of \tsext in terms of a statistical significance is
discussed in section~\ref{monte_carlo_validation}.

The extended source code is speed up in two ways. First, the convolution
of the spatial shape of a radially symmetric extended source with the PSF is
done semi-analytically. The observed photon
distribution can be written as the convolution of the source shape ($I_\text{src}$)
with the psf
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A'.
\end{equation}
The PSF is parameterized by
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $\sigma$ and $\gamma$ have been found empirically as a function
of energy.  For radially symmetric
extended sources, the angular part of the integral can be done analytically
leaving a easier one dimensional integral:
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{equation}
This convolution formula is faster and more closely reduces to the PSF
when the source's size is very small. Decreasing this the numerical
error is important when evaluating the statistical significance of a
source's extension.

The other approximation is to ignore the dependence of the PSF on the
incident angle of a $\gamma$-ray.  Generally, the PSF is a function
of the incident angle of a photon measured with respect to the LAT.
To avoid multiple convolutions of the source shape with the PSF, an
effective PSF is calculated by weighted the PSF for different incident
angles.  Since the convolution has to be done at each iteration of the
extension fit, these approximations were necessary to make the tool run
in a reasonable amount of time.

@@

Errors on fit parameters are typically calculated in one of two
ways. The first method requires explicitly varying one parameter
until the log of the likelihood has fallen by a particular value while
simultaneously optimizing the other fit parameters. The \minuit fitting
package provides the {\em Simplex} function to calculate errors this way
(\cite{minuit_documentation}).  The second method involves estimating the
likelihood function as being a multivariate Gaussian and calculating the
curvature of the log of the likelihood at the peak to estimate when the
function will fall by the same value. \minuit provides the {\em HESSE}
algorithm to calculate errors this way.  The {\em HESSE} errors are
significantly faster but are only approximately true.

We found that the {\em HESSE} algorithm would sometimes badly
underestimate the errors on the extension of the source.  On the other
hand, the {\em Simplex} algorithm was prohibitively slow and would often
fail to convergence. \pointlike's approach to calculating extension errors
is to fixing all the spatial parameters and varying just the extension
until the likelihood has fallen by the desired amount.  This approach
makes the approximation that the covariance between each spatial parameter
is small but not that the likelihood is Gaussian. This method 
is shown schematically in figure~\ref{extension_profile} which shows
the change in log likelihood when varying the fit extension of
the SNR IC443.

The localization error is then calculated by fixing the extension to the
best fit value and fitting an ellipse to the likelihood as a function
of the spatial parameters.  This fit can be used to calculate an
elliptical localization error which correctly accounts for covariance
between the two position coordinates.  This same method is used to
calculate the localization error of all point sources in the LAT catalog
(\cite{first_cat}).

@@

Because of its computational speed, \pointlike is important for analysis
that require many iterations such as source localization and extension
fitting.  On the other hand, we expect \gtlike to get a slightly
more precise answer for fitting the spectral parameters of a source.
Furthermore, because \gtlike is the standard likelihood analysis package,
it has been more extensively validated for spectral analysis.

For that reason, in the following analysis we took the best fit
positions and extensions of the source model fit by \pointlike but
refit the spectral spectral parameters of the model using \gtlike.
\gtlike provides a second calculation of the likelihood which gives a
second measure of \ts and \tsext.

We found very good agreement between the two methods. For the new
extended source candidates, table~\ref{alt_diff_model_results} shows a
comparison of \ts and \tsext computed using \pointlike and \gtlike. For
the rest of this paper, unless otherwise stated all \ts, \tsext, and
fit spectral parameters were found using \gtlike with the localization
results computed by \pointlike.

@@

There is a degeneracy between a spatially extended source and multiple
point sources.  To assess the possibility of source confusion,
we developed a function in \pointlike to simultaneously fit two point
sources and compare the likelihood of two point sources to the likelihood
of one extended source.

The code uses \minuit to simultaneously fit the position of the two point
sources.  To avoid projection effects, the fit is done in a coordinate
system rotated to the celestial equator.  Instead of directly fitting
the rotated coordinates $x_1$ and $x_2$, we decided there would be less
covariance by fitting instead the sum and difference of the rotated
coordinates $S=(x_1+x_2)/2$ and $D=(x_2-x_1)/2$.  The log likelihood
$L(S,D)$ is maximized by fitting the sum and difference of the two point
sources.  Because of \pointlike's computational efficiency, this algorithm
runs in a reasonable amount of time. Similar to extension described in
section~\ref{gtlike_crosscheck}, the spectrum of can be refit using \gtlike
at the best fit positions.

We can define the parameter \tsinc as twice the increase in log likelihood
fitting the region with two point sources compared to fitting the region
with one point source:
\begin{equation}
  \tsinc=2\times(LL_\text{2pts}-LL_\text{ps}).
\end{equation}
A straightforward likelihood ratio test between the two point source
model and the extended source model because the models are note nested.

\tsinc can not be directly compared to \tsext to see which model is better
because the models are not nested (\cite{statistics_with_care}). Even
though the comparison of \tsext with \tsinc is not a calibrated test,
we find the extreme cases where $\tsinc \ll \tsext$ or $\tsinc\gg\tsext$
to be suggestive and use this as a test for source confusion. For the
spatially extended candidates found, we computed both \tsext and \tsinc.
The results are presented in table~\ref{dual_localization_results}

@@

An extended source must be modeled with a spatial shape.
The convolution of that shape with the PSF is fit to the data.
Possible choices for spatial models include a two dimensional Gaussian
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{\pi\sigma^2}\delta\left(x^2+y^2<\sigma^2\right).
\end{equation}
Although these shapes are significantly different, the
convolution of them with the PSF for typical sizes look very
similar.

This is demonstrated in figure~\ref{compare_disk_gauss}.  It shows the
PSF that would be observed for a power-law source of spectral index two,
the probability density function (PDF) for a Gaussian and uniform disk of width
$0.5\deg$, and the PDF for the same Gaussian
and uniform disk convolved with the PSF.  The figure shows the analysis
using an energy range between 1 \gev and 100 \gev and using an energy
range between 10 \gev and 100 \gev.  The convolved shape of the disk and
Gaussian do separate better at higher energy where the PSF is narrower.
This plot shows that the LAT is not very sensitive to probing different
kinds of extended sources.  In this paper we always fit with uniform disk
spatial models.

To get the convolved disk and Gaussian shapes to match in
figure~\ref{compare_disk_gauss}, we had to set them to predict
a 68\% containment radius of 0.5\deg.  The 68\% containment
radius for a spatial model is the radius at which 68\% of the
intensity is enclosed.  It can be computed by a straightforward
integral and we find that $\rsixeight_\text{,disk}=0.82\sigma$
whereas the $\rsixeight_\text{,Gaussian}=1.51\sigma$. Therefore, in
figure~\ref{compare_disk_gauss} the uniform disk has
$\sigma=0.41\deg$ whereas the Gaussian has 
$\sigma=0.75\deg$.

Overlaying fit sizes on top of counts maps can be misleading.
This effect can best be seen in figure~\ref{compare_r68} which shows
an extended source simulated in the energy range from 1 \gev to 100
\gev with a uniform disk spatial model. The simulated source has a size
$\sigma=0.5\deg$, an integral flux between 1 \gev and 100 \gev of $3\times
10^{-8}\ph/\cm^2/\sec$, and a spectral index 2.  To reduce Poisson noise,
the counts map in figure~\ref{compare_r68} is smoothed by a 0.1\deg
Gaussian kernel. Overlaid on the plot is the fit extension assuming the
source has a uniform disk and a Gaussian spatial model.  The solid white
line represents the best fit disk extension $\sigma_\text{disk}=0.496$
The solid black line represents the best fit extension $\sigma$ for
the Gaussian which was found to be $\sigma_\text{Gaussian}=0.2627$.
The white and black dashed lines are the 68\% containment radius of
the fit Disk and Gaussian respectively.  Even though the fit $\sigma$
for the disk and uniform Gaussian seem very different, their 68\%
containment radius are very similar ($\rsixeight_\text{,disk}=0.41$
and $\rsixeight_\text{,Gaussian}=0.40$).

This study shows that the extension $\sigma$ for different spatial
models can be very different even when predicting almost the same PDF.
Since we are not very sensitive to the particular morphology of an
extended source, there is no correct choice for which shape to plot. For
the following analysis, we use a uniform disk as our spatial model
and overlay the disk's edge $\sigma$.  But due to smearing of the PSF,
$\sigma$ typically looks larger than the observed counts distribution.
When comparing the fit extensions with other wavelengths (especially \tev
measurements of galactic sources), it is important to compare consistent
values. If the assumed shapes are different, it is best to indirectly
compare $rsixeight$.

@@

We validated \pointlike for extended source analysis using a series of
Monte Carlo studies.

@@

\tsext can be used to calculate the statistical significance of the
source's extension.  Since the point and extended models are nested, we
would always expect $\tsext>0$.  On the other hand, we would not expect
the likelihood to change very much unless the source was significantly
extended.  Since an extended source has one additional degree of freedom,
Wilks' Theorem says that the distribution of $\tsext$ when testing
a point source for extension should follow a $\chi^2_1$ distribution
(\cite{wilks_theorem}).

On the other hand, since the null hypothesis rests on the edge of
parameter space when the extension is 0, Wilks' Theorem does not hold
(\cite{warn_wilks_theorem}).  One might expect the actual distribution
to be a half $\chi^2_1$ distribution with one degree of freedom plus
half a delta function at 0:
\begin{equation}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
Chernoff proved that under certain assumptions, this is the distribution
that would hold when the null hypothesis lies on the edge of parameter
space (\cite{chernoff}).  The false detection
probability can be estimated by integrating this function from an observed
test statistic value to infinity. It is for this particular distribution
that the often quoted result holds that $\sqrt{\ts}$ is a measure of
the number of $\sigma$ of the detection (\cite{mattox_egret}).

The half $\chi^2_1$ distribution was found by a Monte Carlo study to hold
for the special case of source finding of EGRET sources using a maximum
likelihood analysis (\cite{mattox_egret}).  For half of the tests of a
point source, statistical fluctuations would make the observed counts look
narrower than the point spread function leading to $\tsext=0$. So it is
reasonable to expect that a similar half $\chi^2_1$ would be appropriate
for extension testing. We tested this hypothesis using a dedicated Monte
Carlo study.

Before we show the Monte Carlo simulations, we have to bring up one issue
we found with these tests.  Despite our best efforts, there was always
a small numerical discrepancy between the likelihood when fitting the
source as the PSF and when fitting it as a very small extended source
with This is due to numerical error introduced by the convolution.
In most situations, this error would be insignificant. But it does bias the
distribution of \tsext in the null hypothesis. To correct for this,
instead of computing the likelihood in the null hypothesis by fitting
the PSF, we instead fit the null hypothesis as an extended source with
the extension fixed to be very small.

@@

@@

\begin{table}
  \begin{centering}
    \begin{tabular}{ r | r | r | r }
      \hline
      \hline
      $\gamma$ & flux ($\ph/\cm^2/\sec$) & $N_\text{sims}$ & $\langle\ts\rangle$ \\
      \hline
      1.5 &          $10^{-6}$ &           31952 &  92862 \\
      &  $3\times 10^{-7}$ &           31962 &  22169 \\
      &          $10^{-7}$ &           31977 &   5806 \\
      &  $3\times 10^{-8}$ &           31991 &   1270 \\
      &          $10^{-8}$ &           31940 &    301 \\
      &  $3\times 10^{-9}$ &           30324 &     62 \\
      \hline
      2 &          $10^{-6}$ &           31872 &  22067 \\
      &  $3\times 10^{-7}$ &           31890 &   4898 \\
      &          $10^{-7}$ &           31858 &   1097 \\
      &  $3\times 10^{-8}$ &           31632 &    236 \\
      &          $10^{-8}$ &           27491 &    103 \\
      \hline
      2.5 &          $10^{-6}$ &           31822 &   4706 \\
      &  $3\times 10^{-7}$ &           31822 &    889 \\
      &          $10^{-7}$ &           31169 &    176 \\
      &  $3\times 10^{-8}$ &           21591 &     41 \\
      \hline                                                
      3 &          $10^{-6}$ &           31763 &    929 \\
      &  $3\times 10^{-7}$ &           31665 &    161 \\
      &          $10^{-7}$ &           19271 &     40 \\
      \hline
    \end{tabular}
    \caption{
    This table presents a list of the spectral models (flux and spectral
    index) used in the Monte Carlo study of \pointlike described in
    section~\ref{monte_carlo_validation}.  For each spectral model, the
    number of statistically independent simulations and the average test
    statistic found for that given spectral model ($\langle\ts\rangle$) is tabulated.
    The important feature in this table is that the Monte Carlo
    simulation probes most of the sources parameters that could be detected
    by the LAT.  The simulation probes a wide range of spectral indices
    (from spectral index $\gamma$ of 1.5 to 3) and a wide range of the fluxes
    (varying from a close to $5\sigma$ source detection to a $>10\sigma$
    source detection).  Only simulations which produced statistically
    significant detections ($\ts>25$) were kept.  What is quoted is the
    $>100$ \mev integral flux measured in units of $\ph/\cm^2/\sec$.
    More information about this Monte Carlo analysis is presented in
    section~\ref{monte_carlo_validation}
    }
    \label{ts_ext_num_sims}
  \end{centering}
\end{table}

@@

  \begin{figure}
    \begin{center}
      \includegraphics{mc_plots/compare_disk_gauss.pdf}
      % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/compare_disk_gauss/v2/compare_disk_gauss.pdf
    \end{center}
    \caption{
    This plot demonstrates that different spatial models when smoothed
    out by the PSF look very similar.  The $y$ axis is the log of the
    intensity and the $x$ axis is $\theta^2$.  The black line is the
    PSF that would be observed for a power-law source of spectral index
    two. The dashed red line is the probability density of a Gaussian
    of width $\rsixeight=0.5\deg$. A Gaussian on this plot is just a
    straight line.  The dashed blue line is the probability density for
    a uniform disk of width $\rsixeight=0.5\deg$.  Finally, the solid
    red and blue lines show the convolution of the Gaussian and uniform
    disk respectively with the PSF.  The left plot shows the analysis
    with an energy range from 1 \gev to 100 \gev whereas the right plot
    shows the analysis with an energy range from 10 \gev to 100 \gev.
    }\label{compare_disk_gauss}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics{mc_plots/compare_r68.pdf}
      % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/compare_r68/v2/plot.py
      \end{center}
      \caption{A plot of the smoothed counts for a simulated uniform
        disk extended source in the energy range from 1 \gev to 100 \gev.
        The source has a size $\sigma=0.5\deg$, an integral flux between
        1 \gev and 100 \gev of $3\times 10^{-8}\ph/\cm^2/\sec$, and
        a spectral index 2.  The smoothed counts from a point source
        with the same spectrum are shown in the bottom right inset.
        This source was fit with both a uniform disk and a Gaussian
        spatial model.  The white circle represents the fit size $\sigma$
        fitting with the disk hypothesis.  The black circle is the fit
        $\sigma$ fitting with the Gaussian hypothesis.  The dashed white
        and black circles are the fit \rsixeight defined in section
        \ref{compare_source_size} for the disk and Gaussian hypothesis
        respectively.  This plot shows that fitting the same source with
        different spatial shapes can lead to very different values of
        $\sigma$. On the other hand, the different shapes have a good
        agreement about the 68\% containment radius \rsixeight.
      }\label{compare_r68}
    \end{figure}

>>>>

\pointlike also bins $\gamma$-rays depending upon whether they convert
in the thin or thick layers of the tracker. Since the PSF is smaller for
front entering photons, by binning each conversion type independently,
\pointlike improves the overall sensitivity.  Furthermore, \pointlike
uses a sparse matrix representation of the spatial bins where only the
bins with counts in them are used. To evaluate the likelihood, \pointlike
then independently integrates the model predicted counts in the region.

Intelligent binning and a sparse matrix representation of the data
effectively interpolates between a binned analysis at low energy (where
the resolution is poor and there are many counts) and an unbinned
analysis at high energy (where the resolution is good but there are
few counts). This approach provides significant savings in time while
introducing only small numerical error.

Finally, \pointlike is written mostly in the \python programming
language and uses {\em numpy} and {\em scipy} to vectorize most of the
calculations. This provides a reasonable trade off between computational
efficiency and ease of development. By leaving the interface as a series
of \python modules, advanced functions including extension fitting
(section~\ref{extension_fitting}) and dual localization
(section~\ref{dual_localization_method}) could be quickly
built on top of the core functionality.

@@

We modified \pointlike to allow modeling the spatial extension 
of a source. It is assumed that the spatial and spectral information
completely decouple and that the spatial shape can be represented by a
simple geometric function.  The code consistently convolves the extended
source shape with the PSF (as a function of energy) when computing the
model predicted counts.

\pointlike uses the \minuit fitting library to maximize the likelihood
by simultaneously fitting the position and extension of an extended
source (\cite{minuit_documentation}).  As will be described in
section~\ref{monte_carlo_validation}, this simultaneous fitting
is necessary to correctly calculating the statistical significance
of the detection of extension.
For each position and extension,
the spectral parameters of the sky model are refit. To avoid projection
effects, what is directly fit is not the longitude and latitude of
the source but instead the displacement in a rotated reference frame.

The significance of the extension of a source can be calculated from
\tsext, which is defined as twice the increase in log likelihood between
the extended source model at its best fit values and the point source
model at its best fit values:
\begin{equation}
  \tsext=2\log(LL_\text{ext}/LL_\text{ps}).
\end{equation}
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
section~\ref{monte_carlo_validation}.

The convolution of radially symmetric extended sources is optimized by
using a semi-analytic calculation.  The observed photon distribution
can be written as the convolution of the source shape ($I_\text{src}$)
with the psf
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A'.
\end{equation}
For the LAT, the PSF is parameterized as
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $\sigma$ and $\gamma$ are free parameters.  For radially symmetric
extended sources, the angular part of the integral can be done analytically
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{equation}
This convolution formula is used by \pointlike because it is faster
and more closely reduces to the PSF when the source's size is 
small.

@@

Errors on fit parameters are typically calculated in one of two
ways. The first method requires explicitly varying a parameter
while simultaneously optimizing the other parameters.
until the log of the likelihood has fallen by a particular value.
The \minuit fitting
package provides the {\em Simplex} function to calculate errors this way
(\cite{minuit_documentation}).  The second method involves estimating the
likelihood function as being a multivariate Gaussian and calculating the
curvature of the log of the likelihood at the peak to estimate when the
function will decrease by the desired value. \minuit provides the {\em HESSE}
algorithm to calculate errors this way.  {\em HESSE} errors 
can be calculated significantly faster.

We found that the {\em HESSE} errors on the extension of a source
were sometimes badly estimated.  On the other
hand, the {\em Simplex} algorithm was prohibitively slow and would often
fail to convergence. Our approach to calculating extension errors
is to fix the position of the source and varying just the extension
using the {\em Simplex} algorithm.  This approach
makes the approximation that the covariance between each spatial parameter
is small but not that the likelihood is Gaussian. This method 
is shown schematically in figure~\ref{extension_profile} which shows
the change in log likelihood when varying the extension of
the SNR IC443.
The localization error separately calculated by fixing the extension to the
best fit value and fitting an ellipse to the likelihood as a function
of the spatial parameters. This
elliptical localization error correctly accounts for the covariance
in the position of the source and this same me method is used to
calculate the localization error of all point sources in the first LAT catalog
(\cite{first_cat}).

@@

\pointlike is important for LAT analysis
that require many iterations such as source localization and extension
fitting.  On the other hand, we expect that the spectral paramaters
fit by \gtlike to be slightly more precise.
Furthermore, because \gtlike is the standard likelihood analysis package,
it has been more extensively validated for spectral analysis.
For that reason, in the following analysis we took the best fit
position and extension found by \pointlike and
refit the spectral spectral parameters using \gtlike.
\gtlike also provides a second calculation of the likelihood which gives a
second measure of \ts and \tsext. We do not consider a source to be
significant unless there is agreement between the two methods.

We found very good agreement between the two methods. For the new
extended source candidates, table~\ref{alt_diff_model_results} shows
a comparison of \ts and \tsext computed using \pointlike and \gtlike.
Unless explicitly mentioned, all \ts, \tsext, and spectral parameters
were calculated using \gtlike using \pointlike's positions and extensions.

@@

There is a degeneracy between a spatially extended source and multiple
point sources.  To assess the possibility of source confusion,
we developed a function in \pointlike to simultaneously fit 
the position of two point sources.
The code uses \minuit to fit the position of the two point
sources.  To avoid projection effects, the fit is done in a rotated coordinate
system.  To improve the robustness of the fit, Instead of directly fitting
the rotated coordinates $x_1$ and $x_2$ we instead
fit the sum and difference of the rotated
coordinates $S=(x_1+x_2)/2$ and $D=(x_2-x_1)/2$.  The log likelihood
is then maximized by fitting the sum and difference of the 
sources.  Because of \pointlike's computational efficiency, this algorithm
runs in a reasonable amount of time.

We define \tsinc as twice the increase in log likelihood
fitting the region as two point sources compared to fitting the region
as one point source:
\begin{equation}
  \tsinc=2\times(LL_\text{2pts}-LL_\text{ps}).
\end{equation}
\tsinc can not be directly compared to \tsext to see which model is 
more significant
because the models are not nested (\cite{statistics_with_care}). Even
though the comparison of \tsext with \tsinc is not a calibrated test,
we find the extreme cases $\tsinc \ll \tsext$ or $\tsinc\gg\tsext$
to be suggestive and we only consider a source to be extended if
$\tsext>\tsinc$.

Like extended sources described in section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike. we quote
the spectral values obtained from \gtlike using the best fit positions
found using \pointlike.  For the extended candidates found by our search,
we computed both \tsext and \tsinc.  The results are presented in
table~\ref{dual_localization_results}.

@@

The spatial shape of extended soruces can e modeled
as a two dimensional Gaussian
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}
  \text{PDF}(x,y)=\tfrac{1}{\pi\sigma^2}\delta\left(x^2+y^2<\sigma^2\right).
\end{equation}
Although these shapes are significantly different, the convolution
of them with the PSF for typical sizes look very similar.  Even
though these shapes are significantly different, we demonstrate in
figure~\ref{compare_disk_gauss} that the convolution of shows the
PSF that would be observed for a power-law source of spectral index
2, the probability density function (PDF) for a Gaussian and uniform
disk of width $0.5\deg$, and the convolution of this PDF with the PSF.
The figure shows the convolution for the PSF in the energy range between
1 \gev and and 100 \gev and also in the energy range between 10 \gev and
100 \gev.  The convolved shapesdo separate better at higher energy, but
nevertheless this plot shows that we are not very sensitive to probing
the exact structure of an extended source.  In this paper we always
assume a uniform disk spatial model.

To get the convolved shapes to match in figure~\ref{compare_disk_gauss},
we had to set them to predict the same 68\% containment radius of 0.5\deg.  The
68\% containment radius for a spatial model is the radius at which 68\%
of the intensity is enclosed.  It can be computed by a straightforward
integral and we find that $\rsixeight_\text{,disk}=0.82\sigma$ whereas the
$\rsixeight_\text{,Gaussian}=1.51\sigma$.  Since $\rsixeight=0.5\deg$,
the uniform disk has $\sigma=0.41\deg$ whereas the Gaussian has
$\sigma=0.75\deg$.

This is important when comparing morphologies to other wavelengths
because the quoted size can vary significantly depending on which quanity
is reported.  This effect can be best seen in figure~\ref{compare_r68}
which a counts map of an extended source simulated in the energy
range from 1 \gev to 100 \gev with a uniform disk spatial model and
$\sigma=05$. Overlaid on the plot is the fit extension assuming the
source has a uniform disk and a Gaussian spatial model. The plot overlays
both $\sigma$ and $\rsixeight$ and $\sigma$ is significantly larger for
the disk spatial hypothesis compared to the Gaussian spatial model. On
the other hand, the disk and Gaussian's 68\% containment radius are
very similar.

Since we are not very sensitive to the particular morphology of an
extended source, there is no correct choice for which size to quote
or which shape to overlay. For the following analysis, we quite the
disk's edge and overlay this size.  But due to smearing of the PSF,
$\sigma$ typically looks larger than the observed counts distribution.
When comparing the fit extensions with other wavelengths, it is important
to compare consistent quantities.

@@

@@

We validated \pointlike's extension fitting code with a Monte Carlo
study. \tsext (defined in section~\ref{extension_fitting}) can be used to calculate the statistical significance
of the extension of a source.  Since the point source and extended source models are
nested, an extended source must always fit as well as a point source.  On the other hand, we would
not expect the likelihood to change very much unless the source was
significantly extended.  Since an extended source has one additional
degree of freedom, Wilks' Theorem says that the distribution of $\tsext$
when testing a point source for extension should follow a $\chi^2_1$
distribution (\cite{wilks_theorem}).

On the other hand, since the null hypothesis rests on the edge of
parameter space when the extension is 0, Wilks' Theorem does not hold
(\cite{warn_wilks_theorem}).  One might expect the actual distribution
to be a half $\chi^2_1$ distribution with one degree of freedom plus
half a delta function at 0:
\begin{equation}
  P(\ts)=\tfrac{1}{2}(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
Chernoff proved that under certain assumptions this is the correct distribution
when the null hypothesis is on the edge of parameter
space (\cite{chernoff}).  The false detection
probability can be estimated by integrating this function from an observed
test statistic value to infinity. It is for this particular distribution
that the often quoted result holds that $\sqrt{\ts}$ is a measure of
the number of $\sigma$ of the detection (\cite{mattox_egret}).

The half $\chi^2_1$ distribution was found to hold for the special case
of source finding using EGRET (\cite{mattox_egret}). This distribution
is also plausable when testing point sources for extension. Due to
statistical flucuations, half of the point sources might look narrower
than the PSF leading to $\tsext=0$. The other half of the time, we would
get a distribution consistent with Wilk's theorem. We used a Monte Carlo
study to demonstrat that this is the observed distribution.
 We tested this hypothesis using a dedicated Monte
Carlo study.

@@

Before we present the results of the Monte Carlo simulations, we have to
bring up one issue we found with these tests.  Despite our best efforts,
there was always a small numerical discrepancy between the likelihood when
fitting the source as the PSF and when fitting it as a very small extended
source with This is due to numerical error introduced by the convolution.
In most situations, this error would be insignificant. But it does bias
the distribution of \tsext in the null hypothesis. To correct for this,
instead of computing the likelihood in the null hypothesis by fitting
the PSF, we instead fit the null hypothesis as an extended source with
the extension fixed to be very small.

@@

\begin{figure}
  \begin{center}
    \includegraphics{mc_plots/compare_disk_gauss.pdf}
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/compare_disk_gauss/v2/compare_disk_gauss.pdf
    \end{center}
    \caption{
    This plot demonstrates that different spatial models when smoothed
    by the PSF look very similar.  The $y$ axis is the log of the
    intensity and the $x$ axis is $\theta^2$.  The black line is the
    PSF that would be observed for a power-law source of spectral index
    2. The dashed red line is the probability density of a Gaussian with
    $\rsixeight=0.5\deg$, which is just a straight line on this plot
    The dashed blue line is the probability density for a uniform disk
    with $\rsixeight=0.5\deg$.  Finally, the solid red and blue lines
    show the convolution of the Gaussian and uniform disk with the PSF.
    The left plot shows the analysis with an energy range from 1 \gev to
    100 \gev whereas the right plot shows the analysis with an energy
    range from 10 \gev to 100 \gev.  This plot is discussed futher in
    section~\ref{compare_source_size}
    }\label{compare_disk_gauss}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics{mc_plots/compare_r68.pdf}
      % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/compare_r68/v2/plot.py
      \end{center}
      \caption{
      This plot shows that fitting the same source with different spatial
      shapes can lead to very different values of $\sigma$. On the other
      hand, the different shapes have a good agreement about the 68\%
      containment radius \rsixeight.  It shows counts map for a simulated
      uniform disk extended source in the energy range from 1 \gev to
      100 \gev.  The source has a size $\sigma=0.5\deg$, an integral flux
      between 1 \gev and 100 \gev of $3\times 10^{-8}\ph/\cm^2/\sec$, and
      a spectral index 2.  The smoothed counts from a point source with
      the same spectrum are shown in the bottom right inset.  This source
      was fit with both a uniform disk and a Gaussian spatial model.
      The white circle represents the fit size $\sigma$ modeling the
      source with a uniform disk spatial model.  The black circle is
      the fit $\sigma$ modeling the source with a Gaussian hypothesis.
      The dashed white and black circles are the fit \rsixeight for the
      disk and Gaussian hypothesis. This plot is discussed furhter in
      section~\ref{compare_source_size}
      }\label{compare_r68}
    \end{figure}

    \begin{table}
  \begin{centering}
    \begin{tabular}{ r | r | r | r }
      \hline
      \hline
      $\gamma$ & flux ($\ph/\cm^2/\sec$) & $N_\text{sims}$ & $\langle\ts\rangle$ \\
      \hline
      1.5 &          $10^{-6}$ &           31952 &  92862 \\
      &  $3\times 10^{-7}$ &           31962 &  22169 \\
      &          $10^{-7}$ &           31977 &   5806 \\
      &  $3\times 10^{-8}$ &           31991 &   1270 \\
      &          $10^{-8}$ &           31940 &    301 \\
      &  $3\times 10^{-9}$ &           30324 &     62 \\
      \hline
      2 &          $10^{-6}$ &           31872 &  22067 \\
      &  $3\times 10^{-7}$ &           31890 &   4898 \\
      &          $10^{-7}$ &           31858 &   1097 \\
      &  $3\times 10^{-8}$ &           31632 &    236 \\
      &          $10^{-8}$ &           27491 &    103 \\
      \hline
      2.5 &          $10^{-6}$ &           31822 &   4706 \\
      &  $3\times 10^{-7}$ &           31822 &    889 \\
      &          $10^{-7}$ &           31169 &    176 \\
      &  $3\times 10^{-8}$ &           21591 &     41 \\
      \hline                                                
      3 &          $10^{-6}$ &           31763 &    929 \\
      &  $3\times 10^{-7}$ &           31665 &    161 \\
      &          $10^{-7}$ &           19271 &     40 \\
      \hline
    \end{tabular}
    \caption{
    This table presents a list of the spectral models (flux and
    spectral index) used in the Monte Carlo study of \pointlike
    described in section~\ref{monte_carlo_validation} and presented
    in figure~\ref{ts_ext_mc}.  For each spectral model, the number of
    statistically independent simulations and the average value of \ts
    value ($\langle\ts\rangle$) is tabulated.  The Monte Carlo simulation
    spans parameters representative of sources that could be detected
    by the LAT.  The simulation is of spectral indices $\gamma$ of 1.5,
    2, 2.5, and 3.  The simulation is of fluxes varying from a $5\sigma$
    source detection to a $>10\sigma$ source detection.  Only simulations
    which produced statistically significant detections ($\ts>25$)
    were kept.  What is quoted is the $>100$ \mev integral flux measured
    in units of $\ph/\cm^2/\sec$.  More information about this Monte
    Carlo analysis is presented in section~\ref{monte_carlo_validation}.
    }
    \label{ts_ext_num_sims}
  \end{centering}
\end{table}

@@
